---
title: "Calculating Chi-Squared"
author: "Your Name Here"
output: html_notebook
---

# Molecular Ecology
# Lab 10
## March 14, 2023

### Extra Credit!

Please submit this assignment on gradescope to gain up to 10 extra credit points! 

First, there is an example populations of moths where the chi-squared calculations have been done for you. Then there is a second population, this time of rabbits, where you can did the calculations yourself. All of the numbered sections are opportunities to earn extra credit points. 

So, let's look at the genotypes of some moths and rabbits, find the expected genotypes based on Hardy-Weinberg Equilibrium, and use the Chi-Squared Test to see if the populations are in HWE or not.

1) Read in the data file for the moths
```{r}
moths <- read_csv("path_to_your_file/should/go/here")

#take a look at the data 
moths
```
Calculate the observed genotype frequencies
```{r}
# AA frequency
1469/(1469+138+5)
#or you could type
1469/sum(moths$observations)

# Aa Frequency
138/(1469+138+5)

# aa frequency
5/(1469+138+5)
```

Calculate the allele frequencies
p + q = 1
```{r}
#total number of alleles
(1469+138+5)*2
# total number of A alleles
(1469*2)+138
# total number of a alleles
(5*2)+138

# frequency of A allele
p <- 3076/3224
p

# frequency of a allele
q <- 148/3224
q
```

Calculate the expected genotype frequencies if the population is in Hardy-Weinberg Equilibrium (HWE)
p^2 + 2pq + q^2 = 1
```{r}
p^2

2*p*q

q^2

# we will put these value to a variable that we will use later
freq_values <- c(p^2, 2*p*q, q^2)
freq_values
```

Calculate the expected genotype occurrences in a population of 1612, if the population is in Hardy-Weinberg Equilibrium (HWE)
```{r}
#Expected AA
p^2*1612

#Expected Aa
2*p*q*1612

#Expected aa
q^2*1612

```

Add the expected values to the data table
```{r}
obs_exp <- moths %>% 
  add_column(expected = c(p^2*1612, 2*p*q*1612, q^2*1612)) # c() lets you type a series of things 
  
# take a look at the new column
obs_exp
```

Now let's plot the observed and expected values for the three genotypes
2) Edit the graph to make it unique
```{r}
obs_exp %>% 
  ggplot()+
  geom_col(aes(x = genotype, y = observations), fill = "blue", alpha = 0.8)+
  geom_col(aes(x = genotype, y = expected), fill = NA, color = "red")+
  ggtitle("observations filled in blue, expected outlined in red")
```


Now test if the observed and expected values are significantly different from each other with Chi-Squared Test
We will first do this by typing out the calculations and then with the function in R
```{r}
# first add a new column where the expected value is subtracted from the observed value
obs_exp %>% 
  mutate(obs_minus_exp = observations - expected)

# then add another column where the difference is squared
obs_exp %>% 
  mutate(obs_minus_exp = observations - expected) %>% 
  mutate(diff_sq = obs_minus_exp^2)

# and a final column where the squared value is divided by the expected value
# then add another column where the difference is squared
calc_chi <- obs_exp %>% 
  mutate(obs_minus_exp = observations - expected) %>% 
  mutate(diff_sq = obs_minus_exp^2) %>% 
  mutate(diff_sq_divided_by_exp = diff_sq/expected)

calc_chi

# now we need to sum the values from that final column to get the chi-squared values
sum(calc_chi$diff_sq_divided_by_exp)

# Next let's figure out the degrees of freedom so we can figure out the p-value of our chi-squared test
# degrees of freedom for chi-sq test = (r-1)(c-1) where r is the number of rows and c is the number of columns
# (3-1)*(2-1) = 2*1 = 2 degrees of freedom
```
Now let's enter our chi squared value into this online Chi Square calculator and find the p-value. http://courses.atlas.illinois.edu/spring2016/STAT/STAT200/pchisq.html

Chi-squared test through R function:
```{r}
# we will now use that vector of expected frequencies that we created earlier and used that in the Chi Squared test
# we are telling the function our observed values and the expected frequencies
chisq.test(x = obs_exp$observations, p = freq_values)

```
The null hypothesis is that the observed and expected values are similar and the population is in HWE
The alternative hypothesis is that the observed population is not in HWE

Since our p value is large, we cannot reject the null hypothesis, therefore the population of moths is in HWE



Let's try a second example with rabbits
Read in the data file for the rabbits
```{r}

```

3) Calculate the observed genotype frequencies
```{r}

```

4) Calculate the observed allele frequencies
```{r}


```

5) Calculate the expected genotype frequencies
```{r}

```

6) Calculate the expected genotype occurrences with a population of 8070 and add them to the data table
```{r}

```

7) Now let's plot the observed and expected values for the three genotypes (make it unique! so I know it was your work!)
```{r}

```

8) Calculate the chi-squared value by creating new data columns, like how it was done for the moths
```{r}

```

9) Find the p-value online and type it here:
Is the population in HWE?

10) Calculate the chi-squared value with R
```{r}

```

