---
title: "Calculating R-Squared"
author: "Your Name Here"
output: html_notebook
---
# Molecular Ecology
# Lab 10
# March 14, 2023

First, let's install the packages we will need. This only need to be done once on your computer, then every time after this you can skip this step. 
```{r}
install.packages("tidyverse")
```

Now that we downloaded the package, we need to let R know when we want to use the package
```{r}
library(tidyverse)
```

1. With the package loaded, the next step is to input our data we want to analyze. 
Please add the path to the dataset "fake_h_w_data.csv" that you downloaded from Canvas.
Tip: If you create a folder, save the data file in that folder, and save this R notebook in the same folder, you can just type the name of the file in the "" because your file will be within your working directory. If your data file is not in the same folder as this R notebook, you can type the path to the file. 
```{r}
h_w_data <- read_csv("path_to_your_file/goes_here")

# we have now given the dataset the name "h_w_data" 
```

```{r}
# we can view the data by giving the command for its name:
h_w_data

# take a minute to look a the data set. 
# What are the different columns in the data? How many rows are in the data?
```

2. Next, let's calculate the mean height and weight in our data. The mean height has been done for you, please add the mean weight. 
```{r}
# mean height
mean_height <- mean(h_w_data$height) # the $ is used to tell a function the column you want to use within a data table
mean_height                         #  data_name$column_name

# mean weight
mean_weight <- #fill in the equation for mean weight here
mean_weight
```

Great! We have now looked at our data and found the mean values.

3. Next, we are going to plot the data with x = height, y = weight
We will use "ggplot" to make our figures, it is the most user-friendly way to make figures in R
```{r}
# first, we tell R what data we are working with, then we tell it what we want to do with that data
# %>% this symbol, called "pipe," lets you connect a series of commands, below it is connecting our data frame to ggplot
h_w_data %>% 
  ggplot()+
  geom_point(aes(x =  , y = ))+ # fill in the column names for the variables we want to plot on the graph
  ggtitle("Height and Weight")
```

With this data we are going to run a linear regression, find the best fit line, and then calculate the R^2 and it's p-value.

Next, we want to find the Sum of Squares around the Mean for weight [aka, SS(mean)]
```{r}
# To get the SS(mean), we are going to add a new column to our data that subtracts mean weight (which we just calculated above!) from the weight of each point. 
# the command "mutate" lets you create a new column in your data
hw <- h_w_data %>% 
  mutate(res_mean = weight - mean_weight) %>%  #"res_mean" is the new column with the residuals from the mean
  mutate(res_sq = (res_mean)^2)               #"res_sq" is a new column with the squared residuals
hw #this tells R to show us the new data table we made

# sum of square around the mean
sum_mean <- sum(hw$res_sq)
sum_mean # we now have the sum of square around the mean saved as a variable named "sum_mean"
```

Now let's calculate the variance around the mean, Var(mean)
```{r}
# variance around the mean
var_mean <- sum_mean/length(hw$res_sq) #the length() command will give the length of a column, so the number of rows in a column
var_mean
# or
sum_mean/12  # if you don't want to use the length() command, you can type the number of rows in your data
```

The next step is to find the best fit line for the data. We will do that with a linear regression that uses Least Squares to find the best fit line. 
```{r}
# linear regressions are a type of linear model so the command for a linear regression is "lm"
# the formula for the command is as follows: lm(y_variable ~ x_variable, data = name_of_data_here)
hw_lm <- lm(weight ~ height, data = hw)

summary(hw_lm) # the summary command gives you important information about your linear regression

# y = mx + b
# what is the slope? 
# What is the y-intercept?
# write the formula for the best fit line here:

```

4. Now we will plot the data with the best fit line
```{r}
# To add the regression line to the plot, we need to add a new ggplot command, which is geom_abline(), we will then tell R the y-intercept and the slope of the line we want to add

hw %>% 
  ggplot()+
  geom_point(aes(x = height, y = weight))+
  geom_abline(slope = , intercept = )+ # add the correct values here
  ggtitle("Height and Weight, with line of best fit")
```

5. Now that we have found the best fit line, we can continue to calculate the R^2 for the linear regression of height and weight. We will now create a function that easily give us all the values of the best fit line at all the different x values for our data points. We need this information to complete the Sum of Square around the Best Fit Line, SS(fit).
```{r}
# I have created a function called "y_line" that returns the y value for points along our best fit line at the x values given into the function. 
y_line <- function(x){ 
  y <- <slope>*x + <intercept>  #replace "<slope>" and "<intercept>" with the real values
  return(y)}

hw_fit <- hw %>% 
  mutate(y_fit = y_line(height)) # here we used to function y_line() to fill in the values for the y_fit column that were created. This will add the y values for the best fit line at each height in our dataset. 

hw_fit # take a look at the y_fit column we created to see if the values make sense, to check that our function worked correctly. Compare the values in your hw_fit column to the graph you plotted with the best fit line
```

6. Next, let's get the residuals from our data points to the best fit line. We need this information to calculate the SS(fit). You can do this by following the example I did above with the SS(mean)
```{r}
# sum of square of the fit line
hw_fit2 <- hw_fit %>% 
  mutate(y_res = weight - y_fit) %>% #residuals from the best fit line
  mutate(y_res_sq = y_res^2) # squared residuals

hw_fit2

# now we can find the SS(fit)
# add the command below, it will be similar to what you did for the SS(mean)



```

7. Now let's calculation the Variance around the best fit line, Var(fit)
```{r}
#variance around fit line, Var(fit)
#add the command below to find the variance around the best fit line


```


8. We now have all the components to calculate the R^2 for the linear regression of height and weight
```{r}
# R squared

# (var(mean) - var(fit)) / (var(mean))

# Type the equation for the R^2 with the values you have already computed. 

R_sq <- #type equation here!
R_sq

```

9. Now that you have the R^2 value, does it match the R^2 value that was given by the "lm" output? Scroll up to your linear regression summary and find out if your calculation was correct! It should match the "Multiple R-squared" value in the output from the "lm" command.

10. Next we will calculate the F statistic to find the p-value of the R^2. This will tell us if we have enough data points to support the relationship we found between height and weight.
```{r}
# F_stat = [(SS(mean) -SS(fit))/(pfit - pmean)]/[SS(fit)/(n - pfit)]

#degrees of freedom
# pfit is the number of parameters in our best fit line
# pfit = 

# pmean is the number of parameters in our mean line
# pmean = 

# n is the number of data points we have
# n = 

# pfit - pmean = 
# n - pfit = 

F_stat <- #add equation here
F_stat
    
#Now we will get the p-value tied with the F statistic

pf(F_stat, df = 1, df2 = 10, lower.tail = FALSE) #this will tell you the p-value!

```

Well done! Take a look at the lm output once more. Do your F statistic and p value match the what you calculated? 


11. Now, remake the figure of the data and the best fit line, but this time change the plot in a *unique* way so it is different than your neighbor's graph. You can change the color of the points or line, the shape of the points or the type of line, you can change the background, edit the labels on the axes or the title, and you can change how frequently you have numbers on the axes and the angle of the text! Here is a helpful website that can teach you all about making graphs in R with ggplot. https://r4ds.had.co.nz/data-visualisation.html#data-visualisation

```{r}
#Below are some ideas, but change it however you want it to be! 
#The internet is full of answers, google the things you want to change to learn the code for it!

#you can change a color by typing the color name or by typing the hex code for the color. 
#This website, https://colorpalettes.net/, has color combination suggestions and all the hex codes!

#You can change the shape of your points, and the type of line, examples here: http://www.sthda.com/english/wiki/ggplot2-line-types-how-to-change-line-types-of-a-graph-in-r-software , http://www.sthda.com/english/wiki/ggplot2-point-shapes

# you can also change the background of your figure to have color, no color, lines or no lines: http://www.sthda.com/english/wiki/ggplot2-themes-and-background-colors-the-3-elements 

# Have some fun with it!


#First step, replace the ??? with the missing variables, then get creative!
hw %>% 
  ggplot()+
  geom_point(aes(x =  ??? , y = ????), size = 4, color = "white", fill = "#217074", shape = 23 )+
  geom_abline(intercept = ??? , slope = ????, color = "#8B9D77", linetype = "dashed", size = 1.5)+ 
  labs(title = "Change the title", x = "Height (cm)", y = "Weight (kg)")+
  scale_x_continuous(breaks = seq(0, 140, by = 15), limits = c(0, 140))+
  scale_y_continuous(breaks = seq(0,100, by = 15), limits = c(0,100))+
  coord_fixed()+
  theme(panel.background = element_rect(fill = "#E0F4F5", color = "#E0F4F5"),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45))
```


12. What did you think of this lab? 






